---
title: "RYO Sales Analysis"
author: "Zane Shango"
date: "2/3/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include = FALSE}
library(ggplot2)
library(reshape2)
library(tidyverse)
library(dplyr)


setwd("~/RYO")

# load in data
library(readxl)
ccv17 <- read_excel("2017_CCV_Sales.xlsx")
ccv18 <- read_excel("2018_CCV_Sales.xlsx")
ccv19 <- read_excel("2019_CCV_Sales.xlsx")
ccv_totals <- read_excel("total_ccv_sales.xlsx")



# melt dataframes for plotting using pivor_longer
ccv17_long <- ccv17 %>%
  pivot_longer(
    cols = c('Cigarettes','Vaping','Cigars'), 
    names_to = "Product",
    values_to = "Total_Sold"
  )

ccv18_long <- ccv18 %>%
  pivot_longer(
    cols = c('Cigarettes','Vaping','Cigars'),
    names_to = "Product",
    values_to = "Total_Sold"
  )

ccv19_long <- ccv19 %>%
  pivot_longer(
    cols = c('Cigarettes','Vaping','Cigars'),
    names_to = "Product",
    values_to = "Total_Sold"
  )

ccv_totals_long <- ccv_totals %>%
  pivot_longer(
    cols = c('Cigarettes','Vaping','Cigars'),
    names_to = "Product",
    values_to = "Total_Sold"
  )

# need to set factor levels in month's cronological order for each dataframe
ccv17_long$Month <- factor(ccv17_long$Month, levels = c("JAN", "FEB", "MAR", "APR", "MAY", "JUN", "JUL", "AUG", "SEP", "OCT", "NOV", "DEC"))

ccv18_long$Month <- factor(ccv18_long$Month, levels = c("JAN", "FEB", "MAR", "APR", "MAY", "JUN", "JUL", "AUG", "SEP", "OCT", "NOV", "DEC"))

ccv19_long$Month <- factor(ccv19_long$Month, levels = c("JAN", "FEB", "MAR", "APR", "MAY", "JUN", "JUL", "AUG", "SEP", "OCT", "NOV", "DEC"))

ccv_totals_long$Year <- factor(ccv_totals_long$Year, levels = c("2017","2018","2019"))
```

## 2017 Sales
```{r, echo = FALSE}
ccv17_plot <- ggplot(data = ccv17_long, aes(x = Month, y = Total_Sold, color = Product, group = Product)) +
  geom_point() + 
  geom_line() + 
  ggtitle("2017 Sales") + 
  theme(plot.title = element_text(hjust = 0.5))
ccv17_plot

```

In 2017, the most noticable trend seems to be a spike in purchasing of all products in the month of August. Vaping was noticably increasing in sales almost every month, while Cigarettes stayed mostly stagnant; and Cigars varied a lot on a per month basis, but overall stayed within the same range.

## 2018 Sales
```{r, echo = FALSE}
ccv18_plot <- ggplot(data = ccv18_long, aes(x = Month, y = Total_Sold, color = Product, group = Product)) +
  geom_point() +
  geom_line() + 
  ggtitle("2018 Sales") +
  theme(plot.title = element_text(hjust = 0.5))
ccv18_plot
```

In 2018, Vaping has overtaken Cigars are the top seller, and continues its path of monthly sales growth; however both Cigars and Cigarettes follow the same trend through the year.


## 2019 Sales
```{r, echo = FALSE}
ccv19_plot <- ggplot(data = ccv19_long, aes(x = Month, y = Total_Sold, color = Product, group = Product)) +
  geom_point() + 
  geom_line() + 
  ggtitle("2019 Sales") + 
  theme(plot.title = element_text(hjust = 0.5))
ccv19_plot
```

In 2019, you can see the drastic drop in Vaping after the ban that was put into place at the beginining of September. Cigars and Cigarettes followed a very similar trendline, and both actually experienced a decrease in sales between August and September, but on a smaller scale than Vaping. In 2020, it will be interesting to see how the age increase to purchase these products will affect their sales. 


## Total Analysis
```{r, echo = FALSE}
total_plot <- ggplot(data = ccv_totals_long, 
       aes(x = Year, y = Total_Sold, fill = Product)) +
  geom_bar(stat = 'Identity', position = 'dodge') + 
  ggtitle("2017-2019 Total Sales") + 
  theme(plot.title = element_text(hjust = 0.5))
total_plot
```
```{r, echo = FALSE}
total_faceted <- ggplot(data = ccv_totals_long, 
       aes(x = Product, y = Total_Sold, fill = Product)) +
  geom_bar(stat = 'Identity', position = 'dodge') + 
  facet_wrap(~Year) + 
  ggtitle("2017-2019 Total Sales") + 
  theme(plot.title = element_text(hjust = 0.5))

total_faceted
```

When analyzing total sales, it is clear the trend is shifting more and more to Vaping as the total sales have been climbing over the past 3 years. In fact they've risen from last to first in terms of total sales over this 3 year span. Cigars have been declining each year, but not at a drastic rate; however Cigarettes' incline in sales each year has been enough to overtake Cigars as the second highest selling product. Overall Vaping appears to be the strong trend; but it will be interesting to see how recent law changes regarding tobacco products affect 2020 sales.






