---
title: "Miniproject 3: Mapping COVID Worldwide"
author: "Shahryar Minhas"
date: "3/26/2020"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Guidelines

This homework is to be submitted to the relevant d2l Dropbox folder by 12pm on **April 2**. You may work individually or in a team of at most three people. If you work in a team, please just turn in one assignment and list all names at the top of the assignment. You are allowed to consult any online resource. If you do so, please include a link to the website that you used in your script. If you use code from my lectures, there's no need to include a citation. 

## Data

The only data file for this assignment is `countries-aggregated.csv` and it is up on d2l in the Homework 3 folder. This data set contains the number of confirmed, recovered, and fatal COVID cases by country and time. 

```{r, echo=FALSE}
rm(list=ls())
legis <- read.csv('countries-aggregated.csv', header=TRUE)
```

## Q1

For each country calculate the number of confirmed, recovered and fatal cases of COVID-19 by month. This will leave you with three observations (Jan, Feb, and Mar) by country.

```{r}

```

## Q2

Develop a set of spatial visualizations that show the change in COVID-19 cases by country over time. You should have separate visualizations for the confirmed, recovered, and fatal variables.

```{r}

```

# Q3: Bonus question

Pick any country in the world. Find COVID-19 statistics by some administrative unit (e.g., states in the United States), and develop a spatial visualization to show any COVID related statistics (e.g., number of confirmed cases, number tested).

Further points will be awarded if you are able to show change in number of COVID cases over time.

```{r}

```

